---
title: '1. Introduction'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

This is an `R` markdown notebook, which is a package that comes with RStudio. 
The text above, the part that starts and ends with '---', denotes the front 
matter of your document and can be ignored for now. 

The distinguishing feature of these notebooks is that they contain a mix of 
plain text and `R` code. You can also include LaTex-like code like this: $3+3=6$.
I have dressed up some of my text (with backticks, asterisks, etc.) which changes
the color of some of the text. You will see what effect this has at the end of
this notebook. 

In a notebook, `R` code lives in "code chunks". See lines 26--28 for an example 
of a code chunk. Try executing this code chunk, which adds the number 3 to 
itself, by clicking the *Run* button (the green triangle in the upper right 
of chunk) or by placing your cursor inside it and pressing *Cmd+Shift+Enter*
(*Ctrl+Shift+Enter* on Windows) at the same time. The output should appear
immediately below the chunk as well as in the `R` console. 

```{r}
3+3;
```

If you want to close the output, press the small 'x' at the top right of the 
panel. 

Now, add a new chunk below. To do so, put your cursor on line 38 and click 
*Code* > *Insert Chunk* from the toolbar or  press *Cmd+Option+I* 
(*Ctrl+Alt+I*). Type in your favorite arithmetic operation inside the chunk 
and try running the chunk. 



**Remark 1** `R` does not require that you end each line with a semicolon. 
However, adding one tells `R` that you believe what you have inputted is a
complete piece of code. If `R` disagrees, it will tell you so by throwing an 
error. Therefore, my practice is to use semicolons at the end of each complete 
line of code. 

**Remark 2** I use `R` and `R`Studio interchangeably, because I only ever use 
`R` via `R`Studio. `R`Studio sits "on top" of `R` and makes it easier to use, but
all of the actuall statistical work is done by `R`. 


## Exercises

Read the following. Follow any instructions and answer any questions by 
replacing the `<XXX>` with your response / answer. 

### Packages 

An `R` package is a collection of objects and/or function. Anyone can create 
and share one. All packages fall in one of the following categories:

  (i) installed when you installed `R` and automatically loaded whenever you 
  open `R`Studio
  
  (ii) included with standard `R` installation but requires that you manually
  load the package each time you open a new session of `R`Studio
  
  (iii) not included with standard `R` installation and must be installed (and
  loaded in each new session) before you can use it
  
"Installing" a package means downloading it onto your harddrive and "loading" 
a package means attaching a package's contents into your current session of `R`. 

You only need to install a package once, but you need to load a package anytime 
you open `R`. 
  
The `tidyverse` by Hadley Wickham (winner of the 2019 COPSS award) is one of 
the most popular and revolutionary packages and one that we'll be using today. 
However, it does not come automatically installed with `R`. Run the code chunk 
below to install and then load the `tidyverse` (it may take a minute or two
to download everything). 

```{r}
if(!require(tidyverse)) {
  install.packages("tidyverse");
  library(tidyverse);
}
```

Here's what the code chunk did: `!require()` checks to see if the package is 
already installed. If so, it loads it into your workspace. Otherwise, it will 
install the package and then after installation is complete load it (`require` 
and `library` are roughly synonymous commands). 

Run the above code chunk again. What happened? Why do you think this happened?

<XXX>

### Data

"This study tested the hypotheses that the risk of 30-day mortality associated 
with elective general surgery: 1) increases from morning to evening throughout 
the routine workday; 2) increases from Monday to Friday through the workweek; 
and 3) is more frequent in July and August than during other months of the year.
As a presumed negative control, the investigators also evaluated mortality as a 
function of the phase of the moon. Secondarily, they evaluated these hypotheses 
as they pertain to a composite in-hospital morbidity endpoint."

More information, including a link to the data dictionary, can be found here: https://www.causeweb.org/tshs/surgery-timing/. It will be helpful to have the data dictionary pulled in a browser tab for reference.  


```{r}
surgery <- 
  read_csv(file = "surgery_timing_subset.csv");
```

If your data were read in, you should have received a message that starts 
"Parsed with column specification...". Generally speaking, what is this message 
telling you? 

<XXX>

Although all of the columns except the first have numeric values, we can tell 
from the data dictionary that only a few are actually truly numeric. Let's 
re-read in these data, telling `R` to treat every column as a character except
for the ones we say otherwise:

```{r}
surgery <- 
  read_csv(file = "surgery_timing_subset.csv", 
           col_types = cols(
             #This tells R what the default value 
             #should be for each column, if unspecified
             .default = col_character(),
             #These are the names of the column as written 
             #in the first row of the spreadsheet (or the 
             #second column of the data dictionary):
             age = col_double(), 
             bmi = col_double(), 
             mortality_rsi = col_double(), 
             complication_rsi = col_double(), 
             ccsMort30Rate = col_double(), 
             ccsComplicationRate = col_double(), 
             hour = col_double(), 
             dow = col_integer(),
             month = col_integer()));
```


If you type the name of any object, `R` will interpret that as a print command
and print the object. In this case, typing `surgery;` will print the surgery 
dataset we have just read in. 

```{r}
surgery;
```


### Plots


Now practice making some plots of these data. Run the chunks and try to answer
the questions. 


```{r}
ggplot(data = surgery) + 
  geom_boxplot(mapping = 
                 aes(x = mort30, 
                     y = hour));
```


```{r}
ggplot(data = surgery) + 
  geom_smooth(mapping = 
                 aes(x = hour, 
                     y = as.numeric(complication)));
```


For the above two plots, what 'geom' function was used, and what aesthetics were provided?
Is there anything you would change about either of the plots?

<XXX>

```{r}
ggplot(data = surgery) + 
  geom_bar(mapping = 
               aes(x = month, 
                   fill = complication));
```

```{r}
ggplot(data = surgery) + 
  geom_bar(mapping = 
               aes(x = month, 
                   fill = complication),
           position = "fill");
```

What is the difference between the above two barplots? In other words, what is the
effect of adding 'position = "fill"'?

<XXX>


## Putting it all together

If you want to create a final, finished product of your work to share with your 
partner, children, parents, or grocer, click the *Knit* button or press
*Cmd+Shift+K* (*Ctrl+Shift+K*) to create an HTML file of your entire notebook. 

You should see now the effect of dressing up the text: use *asterisks* for
*italic text*, **double asterisks** for **bold text**, and `single back ticks` 
for `pseudo code`.

If you want to create an PDF or Word file instead, click the down arrow next to 
*Knit* and choose another option; future knits will also knit to this new format
until you change it again. Knitting to PDF may not work if this is your own 
computer and you have never used the LaTeX engine before. Similarly, knitting to 
Word may not work if you do not have Word installed on your computer. 

Another helpful tip: put your cursor anywhere in this document and press 
*Cmd+Option+P* (*Ctrl+Alt+P* in Windows) to run all chunks above the cursor in 
your console. 

You've finished your first notebook! Please indicate as such via the 
polleverywhere poll on the screen. 

## Further reading (if you have time remaining)

You have probably realized by now that most interesting things in `R` happen 
with functions. Read through the chapter from the book 
'R for Data Science' about `R` functions:

https://r4ds.had.co.nz/functions.html#function-arguments

And here is a more fundamental tutorial on functions in `R` from the folks at 
Software Carpentry (you may not have time to read through this all right now):

https://swcarpentry.github.io/r-novice-inflammation/02-func-R/


## References

Wickham, Hadley and Grolemund, Garrett. *R for Data Science*. O'Reilly Media, 
Inc., 2017. https://r4ds.had.co.nz/

