---
title: '2. Tidyverse verbs'
output:
  html_document:
    df_print: paged
  pdf_document: default
---

## Introduction 

In this notebook, you'll hone your tidyverse skills by practicing the use of
some verbs. As before, anything with an `<XXX>` requires an answer from you. 

First, run the chunk immediately below that to load the tidyverse and read in 
the surgery data we've been using:

```{r}
if(!require(tidyverse)) {
  install.packages("tidyverse");
  library(tidyverse);
}
surgery <- 
  read_csv(file = "surgery_timing.subset.csv", 
           col_types = cols(
             #This tells R what the default value 
             #should be for each column, if unspecified
             .default = col_character(),
             #These are the names of the column as written 
             #in the first row of the spreadsheet:
             age = col_double(), 
             bmi = col_double(), 
             mortality_rsi = col_double(), 
             complication_rsi = col_double(), 
             ccsMort30Rate = col_double(), 
             ccsComplicationRate = col_double(), 
             hour = col_double(), 
             dow = col_integer(),
             month = col_integer()));
```

## Demonstration of the verbs

I'll be using the pipe operator, `%>%`, in this document. This is useful when 
iterating functions. For example, instead of `f(g(h(x)))`, I can use the `%>%`
to write: `x %>% h() %>% g() %>% f()`. In words, take `x`, pass it as an 
argument of `f()`, pass its result as an argument to `g()`, and finally pass its
result as an argument to `h()`. At the cost of more whitespace used, I 
can more clearly communicate what operations are being conducted, and in what
order. The link at the very bottom of this notebook points to a helpful page 
explaining in more detail how it works.

Here is what a typical usecase of the `tidyverse` verbs:

```{r}
surgery %>% 
  group_by(baseline_cancer, baseline_cvd) %>%
  summarize(number = n(),
            mean_mortality = mean(mort30 == "1"),
            mean_complication = mean(complication == "1"));
```

In words, I'm taking the `surgery` object, telling `R` to create groups based on
all possible combinations of `baseline_cancer` and `baseline_cvd`, and then, 
within each of those groups, (i) count the number of rows, (ii) count the 
proportion of observations in which the patient died, and (iii) count the 
proportion of observations in which a complication occurred. The mean 
calculations are slightly strange-looking because I have read in the variables
`mort30` and `complication` as character variables and not integers. Thus, I 
am literally counting the proportion of rows in which the variable `mort30` 
equals the character value '1'. 

The following code chunk is identical to the one above but does not use the 
pipe operator `%>%`. 

```{r}
summarize(group_by(surgery, 
                   baseline_cancer, 
                   baseline_cvd), 
          number = n(),
          mean_mortality = mean(mort30 == "1"),
          mean_complication = mean(complication == "1"));
```

It is more complicated to read (or at least that is the argument in favor of 
using `%>%` instead).

## Your turn: interpreting some results

Run the following chunks. Then, using the printed output, your current knowledge
of the `tidyverse`, the data dictionary, and your statistical intuition, 
describe what each of the following chunks does.  

```{r}
surgery %>%
  filter(!is.na(bmi)) %>%
  group_by(hour <= 8) %>%
  summarize(number = n(),
            mean_bmi = mean(bmi),
            mean_mortality = mean(mort30 == "1"),
            mean_complication = mean(complication == "1"));
```

1. `<XXX>`


```{r}
surgery %>%
  filter(baseline_cvd == "1" | 
           baseline_diabetes == "1" |
           baseline_pulmonary == "1") %>%
  summarize(number = n(),
            mean_mortality = mean(mort30 == "1"),
            mean_complication = mean(complication == "1"));
```

2. `<XXX>`


```{r}
surgery %>%
  mutate(complication_rsi_groups = cut(complication_rsi, 
                                       breaks = quantile(complication_rsi), 
                                       include.lowest = T)) %>%
  group_by(complication_rsi_groups) %>%
  summarize(mean_complication = mean(complication == "1"));
```

3. `<XXX>`

```{r}
surgery <- 
  surgery %>%
  mutate(complication_numeric = as.numeric(complication)); 
surgery;
```

4. `<XXX>`

Do any of the above chunks store their results as a new object?

5. `<XXX>`

## Writing your own chunk

Comparing those surgeries that occurred in July or August versus all other 
months, which group had the higher mortality rate? Complete the following chunk
and use the results to answer the question. 

```{r, eval = FALSE}
#This chunk will not run until you've replaced the <XXX> below!
#After you've done so, replace the ```{r, eval = FALSE} above with ```{r}
surgery %>%
  mutate(jul_aug_surgery = month %in% c(7, 8)) %>%
  <XXX>
```

6. `<XXX>`


## Putting it all together

Click the *Knit* button or press *Cmd+Shift+K* (*Ctrl+Shift+K* in Windows)
to create a HTML of your entire notebook that you can share with your favorite 
people.

That's two notebooks completed! Please indicate as such via the polleverywhere
poll on the screen. 

## Further reading (if you have time remaining)

Read through this article explaining how the pipe operator, `%>%`, works,
and why it's recommended:

https://magrittr.tidyverse.org/
